Parallel Computations for Solving Multicriteria Mixed-Integer Optimization Problems

### 2020-12-24 2021/PСT_GergelKozinov/2021_PCT.tex
1

\institute{Institute of Informational Technologies, Mathematics and Mechanics, Mathematical Center, Lobachevsky State University of Nizhny Novgorod, Nizhni Novgorod, Russia \\
\email{gergel@unn.ru, evgeny.kozinov@itmm.unn.ru}
}

@@

Multicriteria optimization (MCO) problems arise almost every time when the optimal options need to be chosen when developing complex technical devices and systems. The scale of demand for MCO problems determines the high intensity of research in this area -- see, for example, monographs \cite{c1,c2,c3,c4,c5} and reviews of scientific and practical results in this area \cite{c6,c7,c8,c9}.

@@

It should be noted that due to the possible changes in the requirements for optimality in the process of calculations it may be necessary to change the parameters of the convolution $\lambda$ from (\ref{eq:8}). Such variations yield a set of scalar global optimization problems (\ref{eq:6})
\begin{equation}\label{eq:10}
\mathbb{F}_T=\{F(\alpha_t,y):1 \leq t \leq T\},
\end{equation}
which is necessary for solving the MCOmix problem. This set of problems can be formed sequentially in the course of calculations; the problems of the set can be solved strictly sequentially or simultaneously in time-shared mode. In addition, the problems of the set $\mathbb{F}_T$ can be solved in parallel using high-performance computing systems. The possibility of forming the set $\mathbb{F}_T$ defines a \textit{new approach to multistage solution of multicriteria optimization} (MMCO) problems -- see, for example, \cite{c22}.

@@

\textit{Rule 3.} Check the stopping condition for the computation. If the required accuracy of the global search is not achieved, it is necessary to supplement the search information $\omega_k$ from (\ref{eq:13}) with the results of the current iteration and continue calculations from Rule 1.

@@

\begin{equation}\label{eq:19}
 \Phi(\alpha,y(x)) = 
\begin{cases}
F(\alpha,y(x),w_1 ),x \in [0,1], \\
F(\alpha,y(x),w_2 ),x \in (1,2], \\
\dots \\
F(\alpha,y(x),w_l ),x \in (l-1,l],
\end{cases}
\end{equation}
where $l$ from (\ref{eq:4}) is the number of different variants of the values of discrete parameters $u \in U$.  The mapping of the extended segment $[0,l]$ onto the domain $D$ of the continuous parameter values from (\ref{eq:2}) can be determined as follows 
\begin{equation}\label{eq:20}
Y(x)=y(x-E(x)),x\in[0,l]
\end{equation}
($E(x)$ denotes the operation of taking the integral part of the number $x$). It should be noted that the function $\Phi(\alpha,y(x))$ is discontinuous at the points $x_i$, $1 \leq i \leq l-1$,  further on, the values of the function $\Phi(\alpha,y(x))$ at these points are considered undefined and not used in the calculations.

@@

At the initial iteration of AGMIS, the value of the minimized function is calculated at some arbitrary point of the interval $(0,l)$(further, obtaining the function value will be called a \textit{trial}). Next, let $k$, $k > 1$, global search iterations be performed. Selection of the trial points $(k+1)$ for the next iteration is determined by the following rules.

@@

To solve each MCOmix problem, 50 uniformly distributed values of convolution coefficients $\lambda \in \Lambda$ from (\ref{eq:8}) were used. To estimate the values of HV and DU efficiency indicators, an approximation of the Pareto domain was calculated for each MCOmix problem  using a uniform grid in the domain of varied values $D\times U$ from (\ref{eq:2})-(\ref{eq:3}). The PAMMIS algorithm was used for the following parameter values: accuracy $\varepsilon=0.05$ from (\ref{eq:26}) and reliability $r=5.6$ (the $r$ parameter is used in PAMMIS to construct estimates of the H{\"o}lder constant $H$ from (\ref{eq:16})).

@@

\begin{table}[ht]
\centering
\caption{Efficiency of the PAMMIS parallel algorithm in solving MCOmix problems}
\label{tab:2}
\begin{tabular}{ccccccc}
\hline
Cores                  & Iterations                  & Speedup                 & DU           & &       & HV                      \\ \hline
\multicolumn{7}{c}{\begin{tabular}[c]{@{}c@{}}The Pareto domain estimate obtained \\ by exhaustive search\end{tabular}}        \\ \hline
48                     & 1 706 667                   & -                       & 20,6         & &       & 29 784,9                \\ \hline
\multicolumn{7}{c}{\begin{tabular}[c]{@{}c@{}}The Pareto    domain  estimate obtained \\ using the PAMMIS method\end{tabular}} \\ \hline
1                      & 86 261,6                    & 1,0                     & 13,8         & &      & 30 212,3                \\
6                      & 15 406,2                    & 5,6                     & 14,4         & &      & 30 317,6                \\
12                     & 6 726,1                     & 12,8                    & 16,2         & &      & 30 248,3                \\
18                     & 5 972,7                     & 14,4                    & 14,3         & &      & 30 551,1                \\
24                     & 3 657,4                     & 23,6                    & 13,9         & &      & 30 092,3                \\
30                     & 3 162,3                     & 27,3                    & 14,9         & &      & 30 443,9                \\
36                     & 2 888,7                     & 29,9                    & 14,6         & &      & 30 528,7                \\
42                     & 2 028,8                     & 42,5                    & 15,6         & &      & 30 046,1                \\
48                     & 1 767,6                     & 48,8                    & 15,3         & &      & 30 255,5                \\ \hline
\end{tabular}
\end{table}

The results of the computational experiments show that the PAMMIS algorithm is scalable as the  speedup of parallel computations increases almost linearly with the number of computing cores used. The obtained values of HV and DU indicators show that the Pareto set estimate using the PAMMIS algorithm is calculated with a greater accuracy and requires a significantly smaller number of trials compared to the results obtained using uniform grids in the domain $D \times U$.

@@

\section*{Acknowledgements} 
This work was supported by the project no. 0729-2020-0055 and by the Research and Education Mathematical Center, project no. 075-02-2020-1483/1

\bibliographystyle{spmpsci}
%\bibliographystyle{unsrt}
\bibliography{mybibfile}

>>>>

\institute{Lobachevsky State University of Nizhny Novgorod, Nizhni Novgorod, Russia \\
\email{gergel@unn.ru, evgeny.kozinov@itmm.unn.ru}
}

@@

Multicriteria optimization (MCO) problems arise almost every time when the optimal decision need to be chosen when developing complex technical devices and systems. The scale of demand for MCO problems determines the high intensity of research in this area -- see, for example, monographs \cite{c1,c2,c3,c4,c5} and reviews of scientific and practical results in this area \cite{c6,c7,c8,c9}.

@@

It should be noted that due to the possible changes in the requirements for optimality in the process of calculations it may be necessary to change the parameters of the convolution $\lambda$ from (\ref{eq:8}). Such variations yield a set of scalar global optimization problems (\ref{eq:6})
\begin{equation}\label{eq:10}
\mathbb{F}_T=\{F(\alpha_t,y , u):1 \leq t \leq T\},
\end{equation}
which is necessary for solving the MCOmix problem. This set of problems can be formed sequentially in the course of calculations; the problems of the set can be solved strictly sequentially or simultaneously in time-shared mode. In addition, the problems of the set $\mathbb{F}_T$ can be solved in parallel using high-performance computing systems. The possibility of forming the set $\mathbb{F}_T$ defines a \textit{new approach to multistage solution of multicriteria optimization} (MMCO) problems -- see, for example, \cite{c22}.

@@

\textit{Rule 3.} Check the stopping condition for the computation. If the required accuracy of the global search is not achieved, it is necessary to supplement the search information $\omega_k$ from (\ref{eq:13}) with the results of the current iteration and continue calculations starting from Rule 1.

@@

\begin{equation}\label{eq:19}
 \Phi(\alpha,y(x)) = 
\begin{cases}
F(\alpha,y(x),w_1 ),x \in [0,1], \\
F(\alpha,y(x),w_2 ),x \in (1,2], \\
\dots \\
F(\alpha,y(x),w_l ),x \in (l-1,l],
\end{cases}
\end{equation}
where $l$ from (\ref{eq:4}) is the number of different variants of the values of discrete parameters $u \in U$.  The mapping of the extended segment $[0,l]$ onto the domain $D$ of the continuous parameter values from (\ref{eq:2}) can be determined as follows 
\begin{equation}\label{eq:20}
Y(x)=y(x-E(x)),x\in[0,l]
\end{equation}
($E(x)$ denotes the operation of taking the integral part of the number $x$). It should be noted that the function $\Phi(\alpha,y(x))$ is discontinuous at the points $x_i$, $1 \leq i \leq l-1$,  further on, the values of the function $\Phi(\alpha,y(x))$ at these points are considered undefined and are not used in the calculations.

@@

At the initial iteration of AGMIS, the value of the minimized function is calculated at some arbitrary point of the interval $(0,l)$(further, obtaining the function value will be called a \textit{trial}). Next, let $k$, $k > 1$, global search iterations be performed. Selection of the trial point $(k+1)$ for the next iteration is determined by the following rules.

@@

To solve each MCOmix problem, 50 uniformly distributed values of convolution coefficients $\lambda \in \Lambda$ from (\ref{eq:8}) were used. To estimate the values of HV and DU efficiency indicators, an approximation of the Pareto domain was calculated for each MCOmix problem  using a uniform grid in the search domain $D\times U$ from (\ref{eq:2})-(\ref{eq:3}). The PAMMIS algorithm was used for the following parameter values: accuracy $\varepsilon=0.05$ from (\ref{eq:26}) and reliability $r=5.6$ (the $r$ parameter is used in PAMMIS to construct estimates of the H{\"o}lder constant $H$ from (\ref{eq:16})).

@@

\begin{table}[ht]
\centering
\caption{Efficiency of the PAMMIS parallel algorithm in solving MCOmix problems}
\label{tab:2}
\begin{tabular}{ccccccc}
\hline
Cores                  & Iterations                  & Speedup                 & DU           & &       & HV                      \\ \hline
\multicolumn{7}{c}{\begin{tabular}[c]{@{}c@{}}The Pareto domain estimate obtained \\ by exhaustive search\end{tabular}}        \\ \hline
48                     & 1706667                   & -                       & 20.6         & &       & 29784.9                \\ \hline
\multicolumn{7}{c}{\begin{tabular}[c]{@{}c@{}}The Pareto    domain  estimate obtained \\ using the PAMMIS method\end{tabular}} \\ \hline
1                      & 86261.6                    & 1.0                     & 13.8         & &      & 30212.3                \\
6                      & 15406.2                    & 5.6                     & 14.4         & &      & 30317.6                \\
12                     & 6726.1                     & 12.8                    & 16.2         & &      & 30248.3                \\
18                     & 5972.7                     & 14.4                    & 14.3         & &      & 30551.1                \\
24                     & 3657.4                     & 23.6                    & 13.9         & &      & 30092.3                \\
30                     & 3162.3                     & 27.3                    & 14.9         & &      & 30443.9                \\
36                     & 2888.7                     & 29.9                    & 14.6         & &      & 30528.7                \\
42                     & 2028.8                     & 42.5                    & 15.6         & &      & 30046.1                \\
48                     & 1767.6                     & 48.8                    & 15.3         & &      & 30255.5                \\ \hline
\end{tabular}
\end{table}

The results of the computational experiments show that the PAMMIS algorithm is scalable as the  speedup of parallel computations increases almost linearly with the number of computing cores used. The obtained values of HV and DU indicators show that the Pareto set estimate using the PAMMIS algorithm is calculated with a greater accuracy and requires a significantly smaller number of trials compared to the results obtained using uniform grids in the search domain $D \times U$.

@@

\section*{Acknowledgements} 
This work was supported by the Ministry of Science and Higher Education of the Russian Federation, project no.0729-2020-0055, and by the Research and Education Mathematical Center, project no. 075-02-2020-1483/1.

\begin{thebibliography}{10}
\providecommand{\url}[1]{{#1}}
\providecommand{\urlprefix}{URL }
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
  \urlstyle{rm}\Url}\fi

\bibitem{c45}
Barkalov, K., Lebedev, I.: Parallel global optimization for non-convex
  mixed-integer problems.
\newblock Communications in Computer and Information Science \textbf{1129},
  98--109 (2019).
\newblock \doi{10.1007/978-3-030-36592-9_9}

\bibitem{c15}
Belotti, P., Lee, J., Liberti, L., Margot, F., W{\" a}chter, A.: Branching and
  bounds tightening techniques for non-convex {MINLP}.
\newblock Optimization Methods and Software \textbf{24}(4-5), 597--634 (2009).
\newblock \doi{10.1080/10556780903087124}

\bibitem{c14}
Boukouvala, F., Misener, R., Floudas, C.: {Global optimization advances in
  mixed-integer nonlinear programming, MINLP, and constrained derivative-free
  optimization, CDFO}.
\newblock European Journal of Operational Research \textbf{252}(3), 701--727
  (2016).
\newblock \doi{10.1016/j.ejor.2015.12.01}

\bibitem{c10}
Branke, J., Deb, K., Miettinen, K., Slowinski, R.: Multiobjective optimization:
  interactive and evolutionary approaches, vol. 5252.
\newblock Springer, Berlin (2008).
\newblock \doi{10.1007/978-3-540-88908-3}

\bibitem{c13}
Burer, S., Letchford, A.: Non-convex mixed-integer nonlinear programming: a
  survey.
\newblock Surveys in Operations Research and Management Science \textbf{17},
  97--106 (2012).
\newblock \doi{10.1016/j.sorms.2012.08.001}

\bibitem{c3}
Collette, Y., Siarry, P.: Multiobjective optimization: principles and case
  studies (decision engineering).
\newblock Springer-Verlag, Berlin, Heidelberg (2011)

\bibitem{c17}
Deep, K., Singh, K., Kansal, M., Mohan, C.: A real coded genetic algorithm for
  solving integer and mixed integer optimization problems.
\newblock Applied Mathematics and Computation \textbf{212}(2), 505--518 (2009).
\newblock \doi{10.1016/j.amc.2009.02.044}

\bibitem{c2}
Ehrgott, M.: Multicriteria optimization.
\newblock Springer-Verlag, Berlin, Heidelberg (2005)

\bibitem{c12}
Eichfelder, G.: Scalarizations for adaptively solving multi-objective
  optimization problems.
\newblock Coldomput. Optim. Appl. \textbf{44}, 249--273 (2009).
\newblock \doi{10.1007/s10589-007-9155-4}

\bibitem{c42}
Evtushenko, Y., Posypkin, M.: Method of non-uniform coverages to solve the
  multicriteria optimization problems with guaranteed accuracy.
\newblock Automation and Remote Control \textbf{75}(6), 1025--1040 (2014).
\newblock \doi{10.1134/S0005117914060046}

\bibitem{c31}
Floudas, C., Pardalos, M.: Recent advances in global optimization.
\newblock Princeton University Press (2016)

\bibitem{c44}
Gaviano, M., Kvasov, D., Lera, D., Sergeyev, Y.: Software for generation of
  classes of test functions with known local and global minima for global
  optimization.
\newblock ACM Transactions on Mathematical Software \textbf{29}(4), 469--480
  (2003)

\bibitem{c19}
Gergel, V.: A unified approach to use of coprocessors of various types for
  solving global optimization problems.
\newblock 2nd International Conference on Mathematics and Computers in Sciences
  and in Industry pp. 13--18 (2015).
\newblock \doi{10.1109/MCSI.2015.18}

\bibitem{c23}
Gergel, V., Barkalov, K., Lebedev, I.: A global optimization algorithm for
  non-convex mixed-integer problems.
\newblock Lecture Notes in Computer Science \textbf{11353}, 78--81 (2019).
\newblock \doi{10.1007/978-3-030-05348-2_7}

\bibitem{c41}
Gergel, V., Barkalov, K., Sysoyev, A.: A novel supercomputer software system
  for solving time-consuming global optimization problems.
\newblock Numerical Algebra, Control \& Optimization \textbf{8}(1), 47--62
  (2018)

\bibitem{c20}
Gergel, V., Kozinov, E.: Accelerating parallel multicriterial optimization
  methods based on intensive using of search information.
\newblock Procedia Computer Science \textbf{108}, 1463--1472 (2017).
\newblock \doi{10.1016/j.procs.2017.05.051}

\bibitem{c21}
Gergel, V., Kozinov, E.: Efficient multicriterial optimization based on
  intensive reuse of search information.
\newblock J. Glob. Optim. \textbf{71}(1), 73--90 (2018).
\newblock \doi{10.1007/s10898-018-0624-3}

\bibitem{c22}
Gergel, V., Kozinov, E.: Multilevel parallel computations for solving
  multistage multicriteria optimization problems.
\newblock Lecture Notes in Computer Science \textbf{12137}, 17--30 (2020)

\bibitem{c7}
Greco, S., Ehrgott, M., Figueira, J.: Multiple criteria decision analysis:
  state of the art surveys.
\newblock Springer (2016).
\newblock \doi{10.1007/978-1-4939-3094-4}

\bibitem{c38}
Grishagin, V., Israfilov, R., Sergeyev, Y.: Comparative efficiency of
  dimensionality reduction schemes in global optimization.
\newblock AIP Conference Proceedings \textbf{1776}, 060011 (2016).
\newblock \doi{10.1063/1.4965345}

\bibitem{c40}
Grishagin, V., Sergeyev, Y., Strongin, R.: Parallel characteristical global
  optimization algorithms.
\newblock Journal of Global Optimization \textbf{10}, 185--206 (1997)

\bibitem{c9}
Hillermeier, C., Jahn, J.: Multiobjective optimization: survey of methods and
  industrial applications.
\newblock Surveys on Mathematics for Industry \textbf{11}, 1--42 (2005)

\bibitem{c36}
Lera, D., Sergeyev, Y.: Lipschitz and {H}{\"o}lder global optimization using
  space-filling curves.
\newblock Applied Numerical Mathematics \textbf{60}(1-2), 115--129 (2010).
\newblock \doi{10.1016/j.apnum.2009.10.004}

\bibitem{c29}
Locatelli, M., Schoen, F.: Global optimization: theory, algorithms, and
  applications.
\newblock {SIAM} (2013)

\bibitem{c6}
Marler, R., Arora, J.: Survey of multi-objective optimization methods for
  engineering.
\newblock Structural and Multidisciplinary Optimization \textbf{26}, 369--395
  (2004).
\newblock \doi{10.1007/s00158-003-0368-6}

\bibitem{c4}
Marler, R., Arora, J.: Multi-objective optimization: concepts and methods for
  engineering (2009)

\bibitem{c1}
Miettinen, K.: Nonlinear multiobjective optimization.
\newblock Springer (1998)

\bibitem{c5}
Pardalos, P., {\v Z}ilinskas, A., {\v Z}ilinskas, J.: Non-convex
  multi-objective optimization, vol. 123.
\newblock Springer (2017)

\bibitem{c30}
Paulavi{\v c}ius, R., {\v Z}ilinskas, J.: Simplicial global optimization.
\newblock Springer-Verlag, New York (2014).
\newblock \doi{10.1007/978-1-4614-9093-7}

\bibitem{c27}
Pint\'er, J.: Global optimization in action (continuous and {Lipschitz}
  optimization: algorithms, implementations and applications).
\newblock Kluwer Academic Publishers, Dordrecht (1996)

\bibitem{c18}
Schl\"{u}ter, M., Egea, J., Banga, J.: Extended ant colony optimization for
  non-convex mixed integer nonlinear programming.
\newblock Comput. Oper. Res. \textbf{36}(7), 2217–2229 (2009).
\newblock \doi{10.1016/j.cor.2008.08.015}

\bibitem{c33}
Sergeyev, Y.: An information global optimization algorithm with local tuning.
\newblock SIAM Journal on Optimization \textbf{5}(4), 858--870 (1995).
\newblock \doi{10.1137/0805041}

\bibitem{c35}
Sergeyev, Y., Famularo, D., Pugliese, P.: Index branch-and-bound algorithm for
  global optimization with multiextremal constraints.
\newblock Journal of Global Optimization \textbf{21}(3), 317--341 (2001).
\newblock \doi{10.1023/A:1012391611462}

\bibitem{c34}
Sergeyev, Y., Grishagin, V.: Parallel asynchronous global search and the nested
  optimization scheme.
\newblock Journal of Computational Analysis and Applications \textbf{3}(2),
  123--145 (2001).
\newblock \doi{10.1023/A:1010185125012}

\bibitem{c37}
Sergeyev, Y., Kvasov, D.: A deterministic global optimization using smooth
  diagonal auxiliary functions.
\newblock Communications in Nonlinear Science and Numerical Simulation
  \textbf{21}(1-3), 99--111 (2015).
\newblock \doi{10.1016/j.cnsns.2014.08.026}

\bibitem{c39}
Sergeyev, Y., Nasso, M., Mukhametzhanov, M., Kvasov, D.: Novel local tuning
  techniques for speeding up one-dimensional algorithms in expensive global
  optimization using {Lipschitz} derivatives.
\newblock Journal of Computational and Applied Mathematics \textbf{383} (2021).
\newblock \doi{10.1016/j.cam.2020.113134}

\bibitem{c25}
Sergeyev, Y., Strongin, R., Lera, D.: Introduction to global optimization
  exploiting space-filling curves.
\newblock Springer, New York (2013)

\bibitem{c32}
Strongin, R., Sergeyev, Y.: Global multidimensional optimization on parallel
  computer.
\newblock Parallel Computing \textbf{18}(11), 1259--1273 (1992).
\newblock \doi{10.1016/0167-8191(92)90069-J}

\bibitem{c24}
Strongin, R., Sergeyev, Y.: Global optimization with non-convex constraints.
  Sequential and parallel algorithms.
\newblock Kluwer Academic Publishers, Dordrecht (2000 (2nd ed. 2013, 3rd ed.
  2014))

\bibitem{c16}
Vigerske, S., Gleixner, A.: {SCIP}: global optimization of mixed-integer
  nonlinear programs in a branch-and-cut framework.
\newblock Optimization Methods and Software \textbf{33}(3), 563--593 (2018).
\newblock \doi{10.1080/10556788.2017.1335312}

\bibitem{c11}
Voutchkov, I., Keane, A.: Multi-objective optimization using surrogates.
\newblock Computational Intelligence in Optimization \textbf{7}, 155--175
  (2010).
\newblock \doi{10.1007/978-3-642-12775-5_7}

\bibitem{c8}
Zavadskas, E., Turskis, Z., Kildiene, S.: State of art surveys of overviews on
  {MCDM}/{MADM} methods.
\newblock Technological and Economic Development of Economy \textbf{20},
  165--179 (2014).
\newblock \doi{10.3846/20294913.2014.892037}

\bibitem{c26}
Zhigljavsky, A.: Theory of global random search.
\newblock Kluwer Academic Publishers, Dordrecht (1991)

\bibitem{c28}
Zhigljavsky, A., {\v Z}ilinskas, A.: Stochastic global optimization, vol.~9.
\newblock Springer, Berlin (2008).
\newblock \doi{10.1007/978-0-387-74740-8}

\bibitem{c43}
{\v Z}ilinskas, A., {\v Z}ilinskas, J.: Adaptation of a one-step worst-case
  optimal univariate algorithm of bi-objective {Lipschitz} optimization to
  multidimensional problems.
\newblock Commun Nonlinear Sci Numer Simulat \textbf{21}(1-3), 89--98 (2015).
\newblock \doi{10.1016/j.cnsns.2014.08.025}

\end{thebibliography}
